NumLin: Linear Types for Linear Algebra

### 2019-01-10 write-up/paper/discussion.tex
Update to LIPIcs style

Using linear types for BLAS routines is a particularly good domain fit
(given the implicit restrictions on aliasing arguments), and as a
result the idea of using substructural types to express array
computations is not a particularly new
one\cite{scholz,henriksen,bernardy2016}.  However, many of these
designs have been focused on building languages to \emph{implement}
the kernel linear algebra functions, and as a result, they tend to add
additional limitations on the language design. Both
Futhark~\cite{henriksen} and Single Assignment C~\cite{scholz} omit
higher-order functions facilitating compilation to GPUs. The work of
\citet{bernardy2016} forbids term-level recursion, in order to ensure
that all higher-order computations can be statically normalized away
and thereby maximize opportunities for array fusion.

@@

This approach is similar to the one taken in linear algebra libraries
for Rust -- these libraries typically take advantage of the
distinction that Rust's type system offers between mutable
views/references to arrays.  The work of \citet{weiss} and
\citet{rustbelt} suggest that Rust's borrow-checker can be expressed
in simpler terms using \emph{fractional-permissions}, though to our
knowledge the programmer-visible lifetime analysis in Rust has never
been formalized.

Using fractional permissions directly, as we did, rather than relying
on a complex has two main benefits. First, our type system
demonstrates that type systems for fractional permissions can be
\emph{dramatically} simpler than existing state-of-the-art approaches,
including both industrial languages like Rust, as well as the academic
languages developed by \citet{bierhoff}.  Their type system, much like
Rust, build a complex dataflow analysis into the typing rules to infer
when variables can be shared or not. This allows for very
natural-looking user programs, but creates the impression that using
fractional permissions requires a heavy theoretical and engineering
effort going well beyond that needed for supporting basic linear
types.

>>>>

Using linear types for BLAS routines is a particularly good domain fit
(given the implicit restrictions on aliasing arguments), and as a
result the idea of using substructural types to express array
computations is not a particularly new
one\cite{scholz,henriksen,bernardy2016}.  However, many of these
designs have been focused on building languages to \emph{implement}
the kernel linear algebra functions, and as a result, they tend to add
additional limitations on the language design. Both
Futhark~\cite{henriksen} and Single Assignment C~\cite{scholz} omit
higher-order functions facilitating compilation to GPUs. The work of
\cite{bernardy2016} forbids term-level recursion, in order to ensure
that all higher-order computations can be statically normalized away
and thereby maximize opportunities for array fusion.

@@

This approach is similar to the one taken in linear algebra libraries
for Rust -- these libraries typically take advantage of the
distinction that Rust's type system offers between mutable
views/references to arrays.  The work of \cite{weiss} and
\cite{rustbelt} suggest that Rust's borrow-checker can be expressed
in simpler terms using \emph{fractional-permissions}, though to our
knowledge the programmer-visible lifetime analysis in Rust has never
been formalized.

Using fractional permissions directly, as we did, rather than relying
on a complex has two main benefits. First, our type system
demonstrates that type systems for fractional permissions can be
\emph{dramatically} simpler than existing state-of-the-art approaches,
including both industrial languages like Rust, as well as the academic
languages developed by \cite{bierhoff}.  Their type system, much like
Rust, build a complex dataflow analysis into the typing rules to infer
when variables can be shared or not. This allows for very
natural-looking user programs, but creates the impression that using
fractional permissions requires a heavy theoretical and engineering
effort going well beyond that needed for supporting basic linear
types.

### 2019-01-10 write-up/paper/paper.tex
Update to LIPIcs style

\documentclass[runningheads,a4paper,twoside]{llncs}
\usepackage{hyperref}
\hypersetup{%
    colorlinks=true,
    linkcolor={black},
    citecolor={black},
    urlcolor={blue}
}

@@

% Ott Rules
\usepackage{supertabular}%
\usepackage[margin=1in]{geometry}%
\usepackage{ifthen}%
\usepackage{alltt}%hack%

@@

@@

% \titlerunning{NumLin: Linear Types for Linear Algebra}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here

\newcommand{\lang}{\textsc{NumLin}}

\author{Dhruv C.~Makwana\inst{1}\orcidID{orcidID} \and
Neelakantan R.~Krishnaswami\inst{2}\orcidID{orcidID}}
%
\authorrunning{D.C. Makwana and N.R. Krishnaswami}

\institute{\email{dcm41@cam.ac.uk}
\url{dhruvmakwana.com} \and
Department of Computer Science, University of Cambridge\\
\email{nk480@cl.cam.ac.uk}}
%
\maketitle
%
\begin{abstract}
    We present \lang, a functional programming language designed to express
    the APIs of low-level linear algebra libraries (such as BLAS/LAPACK) safely
    and explicitly, through a brief description of its key features and several
    illustrative examples. We show that \lang's type system is sound and that
    its implementation improves upon na{\"i}ve implementations of linear
    algebra programs, almost towards C-levels of performance. Lastly, we contrast
    it to other recent developments in linear types and show that using linear
    types and fractional permissions to express the APIs of low-level linear
    algebra libraries is a simple and effective idea.
\keywords{numerical, linear, algebra, types, permissions, OCaml}
\end{abstract}

@@

\bibliographystyle{splncs04}

\bibliography{ourbib}

>>>>

\documentclass[a4paper,UKenglish]{lipics-v2019}
%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"

\bibliographystyle{plainurl}

\author{Dhruv C.~Makwana}{Unaffiliated \url{dhruvmakwana.com} }{dcm41@cam.ac.uk}{[orcid]}{}

\author{Neelakantan R.~Krishnaswami}{Department of Computer Science and Technology, University of Cambridge, United Kingdom}{nk480@cl.cam.ac.uk}{[orcid]}{}

\authorrunning{D.\,C. Makwana and N.\,R. Krishnaswami}

\Copyright{Dhruv C. Makwana and Neelakantan R. Krishnaswami}% LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[300]{Theory of computation~Program specifications}

\keywords{numerical, linear, algebra, types, permissions, OCaml}

\supplement{\url{www.github.com/dc-mak/lt4la}}

%\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}

@@

% Ott Rules
\usepackage{supertabular}%
% \usepackage{geometry}%
\usepackage{ifthen}%
\usepackage{alltt}%hack%

@@

% NumLin
\newcommand{\lang}{\textsc{NumLin}}

@@

\begin{abstract}
    We present \lang, a functional programming language designed to express
    the APIs of low-level linear algebra libraries (such as BLAS/LAPACK) safely
    and explicitly, through a brief description of its key features and several
    illustrative examples. We show that \lang's type system is sound and that
    its implementation improves upon na{\"i}ve implementations of linear
    algebra programs, almost towards C-levels of performance. Lastly, we contrast
    it to other recent developments in linear types and show that using linear
    types and fractional permissions to express the APIs of low-level linear
    algebra libraries is a simple and effective idea.
\end{abstract}

@@

\clearpage
\bibliography{ourbib}
